Hierarchical wrinkling in a confined permeable biogel

### 2015-02-25 fig_plis/article_plis.tex
documentclass sur une seule ligne pour être compatible avec texcount
PNAS version with figures at the end

\documentclass[twocolumn,superscriptaddress,showpacs,preprintnumbers,
amsmath,amssymb,prl]{revtex4-1}
\usepackage[utf8]{inputenc}

\usepackage{graphicx} 
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage[range-units = single]{siunitx}
%\usepackage{multirow}
\usepackage{bm}
%\usepackage{dcolumn,ulem}
\usepackage{amsfonts}

@@

\begin{document}
\title{Hierarchical wrinkling in a confined permeable biogel}
\author{Mathieu Leocmach}
\affiliation{Universit\'e de Lyon, Laboratoire de Physique, \'Ecole Normale Sup\'erieure de Lyon, CNRS UMR 5672, 46 All\'ee d'Italie, 69364 Lyon cedex 07, France}
\author{Mathieu Nespoulous}
\affiliation{Aix-Marseille Université, CNRS, MADIREL UMR 7246, Marseille, France}
\author{Sébastien Manneville}
\affiliation{Universit\'e de Lyon, Laboratoire de Physique, \'Ecole Normale Sup\'erieure de Lyon, CNRS UMR 5672, 46 All\'ee d'Italie, 69364 Lyon cedex 07, France}
\author{Thomas Gibaud}
\email{thomas.gibaud@ens-lyon.fr}
\affiliation{Universit\'e de Lyon, Laboratoire de Physique, \'Ecole Normale Sup\'erieure de Lyon, CNRS UMR 5672, 46 All\'ee d'Italie, 69364 Lyon cedex 07, France}

\begin{abstract}
The slow acidification of caseins, a family of milk proteins, is well known to result in the formation of gels, e.g. yoghurt. Less known is the partial redissolution of the gel at even lower pH. Here, using caseins as pH-responsive building blocks and the acidulent glucono-$\delta$-lactone, we design a model biogel film that spontaneously wrinkles in cascade, provided non-adhesive confinement. Using a combination of titration, rheology, light microscopy and confocal microscopy, we demonstrate that, during continuous acidification, the gel first shrinks and then swells, inducing wrinkling. Taking into account both Darcy flow through the gel and Poiseuille flow in the surrounding solvent, we develop a model that predicts correctly the wrinkling wavelength. Our results should be universal for acid-induced protein gels as they are based on pH-induced charge stabilisation/destabilisation. As such our results could set a benchmark to gain fundamental insights into wrinkled biological tissues, to texture food or to design surfaces for optical purposes.
\end{abstract}

@@

\section*{Introduction}

%There are many ways and reasons for a film to wrinkle. An elastic film buckles due to excess area with respect to its boundaries. Wrinkling is a buckling hindered by a substrate: the mismatch of elastic properties between the film and the substrate selects the preferred wrinkling wavelength~\cite{Gough1940, Bijlaard1946}. On the engineering side the last two decades have seen a bloom of methods to obtain well-controlled patterns via linear~\cite{Bowden1998, Genzer2006, Hu1998, Kim2010, Vandeparre2011, Li2013} or nonlinear wrinkling~\cite{Efimenko2005, Guvendiren2010, Kim2011a, Brau2011}. Such patterns can be triggered by temperature dilation, swelling or the removal of pre-strain. On the biology side wrinkling-controlled morphogenesis is ubiquitous: ageing and the loss of elastic fibres makes our skin wrinkles~\cite{Bissett1987, Genzer2006}, difference in growth rates between the gut tube and its dorsal anchoring is responsible for the vilification of guts~\cite{Savin2011, Ciarletta2014, Shyer2013}, localized cell death in biofilms focuses mechanical forces and initiates 3D labyrinth pattern~\cite{Trejo2013, Asally2012}. On the physics side analogues of wrinkling have been found in situations where an effective substrate comes from the resistance to uniaxial stretching~\cite{Cerda2003}, from gravity~\cite{Kolinski2009, Vella2009, Pineirua2013, Lucantonio2013}, from boundaries~\cite{Vandeparre2011a, Li2013} or from adhesion~\cite{Vella2009a}. Dealing only with (semi)infinite substrates, Biot generalised the elasticity of both the film and the substrate to viscoelasticity~\cite{Biot1957} or poroelasticity~\cite{Biot1964}, leading to time-dependent kinetic wavelength selection. Stability analysis of a film lying on a thin viscous substrate was studied only recently~\cite{Huang2002}. Since in practice gravity sets the wavelength in such situation~\cite{Smoluchowski1910}, the wrinkling on a confined viscous substrate has not been studied experimentally.

There are many ways and reasons for a film to wrinkle. An elastic film buckles due to excess area with respect to its boundaries and wrinkles when buckling is hindered by a substrate. The selection of the preferred wrinkling wavelength is peculiar to the situation and is related to the mismatch of elastic properties between the surface and the substrate~\cite{Gough1940, Bijlaard1946, Biot1957, Bowden1998, Cerda2003, Genzer2006}. On the biology side wrinkling-controlled morphogenesis is ubiquitous. Ageing and the loss of elastic fibres makes our skin wrinkle~\cite{Bissett1987, Genzer2006}. Difference in growth rates between the gut tube and its dorsal anchoring is responsible for the vilification of guts~\cite{Savin2011, Ciarletta2014, Shyer2013}. Localised cell death in biofilms focuses mechanical forces and initiates 3D labyrinth pattern~\cite{Trejo2013, Asally2012}. On the physics side the last two decades have seen a bloom of methods to obtain well-controlled patterns via linear~\cite{Bowden1998, Genzer2006, Hu1998, Kim2010, Vandeparre2011, Li2013} or nonlinear wrinkling~\cite{Efimenko2005, Guvendiren2010, Kim2011a, Brau2011}. Such patterns can be triggered by temperature dilation~\cite{Bowden1998}, swelling~\cite{Hu1998,Kim2010} or the removal of pre-strain~\cite{Genzer2006}.

%ph swelling: hu_1998_Polymer gels with engineered environmentally and kim_2010_Dynamic display of biomolecular patterns
%compressive stress: kim_2011_Hierarchical folding of elastic membranes under

@@

In the following we first describe the physical chemistry that allows us to form such patterns.
We then show how transverse confinement leads to saturation of the amplitude on both the top and bottom walls followed by a cascade of secondary and ternary patterns. Finally we model the situation and identify two possible selection mechanisms, namely porous Darcy flow or viscous Poiseuille flow. Comparing our experimental data to these two models we show that a combination of both Darcy and Poiseuille mechanisms accurately describes our data. This combined model pins down the dynamical origin of the constraints exerted on the gel and nicely predicts the wrinkling wavelength.
%Poiseuille~\cite{Poiseuille1842} flows of the solvent above and below the gel film for long wavelengths and a new mechanism, Darcy~\cite{Darcy1856} flows of the solvent through the gel film for short wavelengths.

\section*{System design}

@@

%On the Stability of Casein Micelles, Pieter Walstra J. of dairy science 1990

@@

%XXXCe paragraphe fait-il sens ici ?XXX As seen on the confocal measurements, the gel is porous and is permeable to the solvent.  This property is determinant as it allow the gel to absorb and expel the solvent and therefore change its volume when sticky boundary are turned off. To quantify this effect, we make the casein gel inside a capillary and dip vertically the capillary in an aqueous buffer~\cite{VanDijk1986}. In the Supplementary Figure~\ref{S-fig:permeability}, we observe that the aqueous buffer goes through the gel and rises in the tube due to capillary forces. Exploitation of this experiment give the permeability of the gel: \SI{67000}{\square\nano\metre}.

@@

%We coat the glass slides with acrylamide brushes to prevent the adhesion of the casein. In fig.2c, when the top slide of the optical cell is coated with acrylamide brushes, we then observe that the casein network detaches from the top slide. $\xi$ does not peak at the isoelectric point but remains constant

%Boundary conditions is the final parameter that is essential to form wrinkled patterns. Once the space spanning network is formed and sticks to the container, protein attraction can only proceed by local shrinking of the protein agregates, leading to larger pores.

%The difference in value of $\xi$ between the fully un-coated  and the half-coated optical cell demonstrate that the gel is subject to constrains as the pH decreases. The relaxation of thoses constrain is key to the wrinkling formation when both the top and bottom slide are coated with acrylamide brushes.

%on a change les condition daccroche ce aui permet au contrainte de se relacher

%The slow acidification of a protein (sodium caseinate) solution down to the isoelectric pH induces gelation~\cite{Bremer1989} (Fig.~\ref{fig:acidification}a). Since caseins strongly adsorb to interfaces, the resulting gel sticks to the container walls, allowing rheological measurements (Fig.~\ref{fig:acidification}b). Further decrease of the pH beyond the isoelectric point makes the casein aggregates swell (Supplementary Fig.~\ref{S-fig:nonmonotonic}), thus weakening the gel~\cite{Braga2006}. In a microscopy cell of slit geometry coated so that the casein cannot adsorb on the top wall, this microscopic nonmonotonic behaviour translates into little variation of the pore size $\xi$ but to solvent expulsion out of the gel phase (syn\ae{}resis) followed by global swelling (Fig.~\ref{fig:acidification}c and Supplementary Fig.~\ref{S-fig:nonmonotonic}).

\section*{Dynamics of the wrinkling process}

@@

\section*{Models and scaling laws}

%Longitudinally constrained swelling of the gel film is the origin and driving force of the buckling instability. The gel area becomes in excess compared to the available area, thus the gel must fold. However, this does not explains how a wavelength is selected.

@@

%L. Landau and E.M. Lifshitz, Theory of Elasticity (Pergamon, New York, 1986), 3rd ed.

@@

@@

\section*{Experimental tests}

@@

\section*{Conclusion}

@@

%role of gravity

\section*{Methods summary}

@@

3D data are collected on a Zeiss LSM510 confocal microscope using \SI{532}{\nano\meter} laser excitation. We use either a 10x (air) lens to measure the geometry of wrinkling, see Supplementary Material, or a 63x (oil) lens to observe the microstructure of the gel. Larger scope pictures, obtained either by stitching fluorescent microscopy images (Nikon Eclipse Ti) or by transmitted/reflected light macroscope (Nikon SMZ745T/Leica DMS1000) are used to measure the wavelength of the primary pattern, Supplementary Figures~\ref{S-fig:generations} and \ref{S-fig:patterns}. The secondary pattern wavelength is taken as half of the primary wavelength, according to Supplementary Figure~\ref{S-fig:generations}.
Data shown in Figure~\ref{fig:acidification}a and c are obtained with a SevenCompact pH-meter (Mettler Toledo) and an MCR 301 rheometer (Anton Paar). For every gel composition we monitor the gelation by small amplitude oscillations (strain 0.1\%, frequency \SI{1}{\hertz}) and estimate the flexural modulus $B$ from the maximum value of the shear modulus $G^\prime$ as $B=2G^\prime/(1-\nu)$ taking $\nu=0.3$ for the Poisson ratio, a value typical of a spongy network~\cite{Greaves2011}.
Permeability is measured using a protocol found in \cite{VanDijk1986}, see Supplementary Method and Supplementary Figure~\ref{S-fig:permeability}. The amount of free casein protein present in the gel during the GDL acidification process is measured by titration and UV-spectrometry (Ocean Optics) according to the protocol found in~\cite{Roefs1986} and detailed in Supplementary Method.

\bibliographystyle{naturemag3}
\bibliography{Yaourt}

\paragraph*{Acknowledgements}
Confocal experiments were conducted at SFR BioSciences Gerland - Lyon Sud (US8 / UMS3444).
The authors thank Jos\'e Bico, Arezeki Boudaoud, Cyprien Gay, L. Mahadevan and Olivier Pierre-Louis for theoretical insights at various stages of the research. 
Precious experimental ideas came from Catherine Barentin, Denis Bartolo, Elisabeth Bouchaud and Thibaut Divoux.
Alan Parker at Firmenich provided the casein and GDL. 
Special thanks to Madame Biot who made M.A. Biot work accessible through the \href{http://www.olemiss.edu/sciencenet/poronet/}{poronet} website. 
ML thanks the Region Rhône Alpes and the Programme d'Avenir Lyon - Saint Etienne (PALSE NoGELPo) for postoctoral grant. 
TG and MN acknowledge funding from the Agence Nationale de la Recherche (ANR-11-PDOC-027). 
SM and ML acknowledge funding from the European Research Council under the European Union's Seventh Framework Program (FP7/2007-2013) / ERC grant agreement No. 258803 and from Institut Universitaire de France.

\paragraph*{Author Contributions}
ML and MN conducted preliminary experiments. ML, SM and TG designed the experiments and interpreted the results. ML and TG performed quantitative experiments. ML performed data analysis and modelling. ML, SM and TG wrote the manuscript.

\paragraph*{Author Information} 
The authors declare that they have no competing financial interests. 
Correspondence and requests for materials should be addressed to TG (\href{mailto:thomas.gibaud@ens-lyon.fr}{thomas.gibaud@ens-lyon.fr}).

>>>>

%% PNAStwoS.tex
%% Sample file to use for PNAS articles prepared in LaTeX
%% For two column PNAS articles
%% Version1: Apr 15, 2008
%% Version2: Oct 04, 2013

%% BASIC CLASS FILE
\documentclass{pnastwo}
\usepackage[utf8]{inputenc}

\usepackage{graphicx} 
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage[range-units = single]{siunitx}
%\usepackage{multirow}
%\usepackage{bm}
%\usepackage{dcolumn,ulem}
%\usepackage{amsfonts}

@@

\begin{document}
\title{Hierarchical wrinkling in a confined permeable biogel}
\author{Mathieu Leocmach\affil{1}{Universit\'e de Lyon, Laboratoire de Physique, \'Ecole Normale Sup\'erieure de Lyon, CNRS UMR 5672, 46 All\'ee d'Italie, 69364 Lyon cedex 07, France},
Mathieu Nespoulous\affil{2}{Aix-Marseille Université, CNRS, MADIREL UMR 7246, Marseille, France},
Sébastien Manneville\affil{1}{},
Thomas Gibaud\affil{1}{}}
%\email{thomas.gibaud@ens-lyon.fr}

\significancetext{ML and MN conducted preliminary experiments. ML, SM and TG designed the experiments and interpreted the results. ML and TG performed quantitative experiments. ML performed data analysis and modelling. ML, SM and TG wrote the manuscript.}

@@

\begin{article}
\begin{abstract}
The slow acidification of caseins, a family of milk proteins, is well known to result in the formation of gels, e.g. yoghurt. Less known is the partial redissolution of the gel at even lower pH. Here, using caseins as pH-responsive building blocks and the acidulent glucono-$\delta$-lactone, we design a model biogel film that spontaneously wrinkles in cascade, provided non-adhesive confinement. Using a combination of titration, rheology, light microscopy and confocal microscopy, we demonstrate that, during continuous acidification, the gel first shrinks and then swells, inducing wrinkling. Taking into account both Darcy flow through the gel and Poiseuille flow in the surrounding solvent, we develop a model that predicts correctly the wrinkling wavelength. Our results should be universal for acid-induced protein gels as they are based on pH-induced charge stabilisation/destabilisation. As such our results could set a benchmark to gain fundamental insights into wrinkled biological tissues, to texture food or to design surfaces for optical purposes.
\end{abstract}

\keywords{wrinkling | porosity | gel | syn\ae{}resis}

\abbreviations{GDL, glucono-$\delta$-lactone}

\dropcap{T}here are many ways and reasons for a film to wrinkle. An elastic film buckles due to excess area with respect to its boundaries and wrinkles when buckling is hindered by a substrate. The selection of the preferred wrinkling wavelength is peculiar to the situation and is related to the mismatch of elastic properties between the surface and the substrate~\cite{Gough1940, Bijlaard1946, Biot1957, Bowden1998, Cerda2003, Genzer2006}. On the biology side wrinkling-controlled morphogenesis is ubiquitous. Ageing and the loss of elastic fibres makes our skin wrinkle~\cite{Bissett1987, Genzer2006}. Difference in growth rates between the gut tube and its dorsal anchoring is responsible for the vilification of guts~\cite{Savin2011, Ciarletta2014, Shyer2013}. Localised cell death in biofilms focuses mechanical forces and initiates 3D labyrinth pattern~\cite{Trejo2013, Asally2012}. On the physics side the last two decades have seen a bloom of methods to obtain well-controlled patterns via linear~\cite{Bowden1998, Genzer2006, Hu1998, Kim2010, Vandeparre2011, Li2013} or nonlinear wrinkling~\cite{Efimenko2005, Guvendiren2010, Kim2011a, Brau2011}. Such patterns can be triggered by temperature dilation~\cite{Bowden1998}, swelling~\cite{Hu1998,Kim2010} or the removal of pre-strain~\cite{Genzer2006}.

@@

%fig:dynamics

In the following we first describe the physical chemistry that allows us to form such patterns.
We then show how transverse confinement leads to saturation of the amplitude on both the top and bottom walls followed by a cascade of secondary and ternary patterns. Finally we model the situation and identify two possible selection mechanisms, namely porous Darcy flow or viscous Poiseuille flow. Comparing our experimental data to these two models we show that a combination of both Darcy and Poiseuille mechanisms accurately describes our data. This combined model pins down the dynamical origin of the constraints exerted on the gel and nicely predicts the wrinkling wavelength.

\section{System design}

%fig:acidification

@@

@@

@@

\section{Dynamics of the wrinkling process}

@@

%fig:sideview

\section{Models and scaling laws}

@@

@@

%fig:DarcyPoiseuille

@@

\section{Experimental tests}

@@

\section{Conclusion}

@@

\begin{materials}

@@

3D data are collected on a Zeiss LSM510 confocal microscope using \SI{532}{\nano\meter} laser excitation. We use either a 10x (air) lens to measure the geometry of wrinkling, see Supplementary Material, or a 63x (oil) lens to observe the microstructure of the gel. Larger scope pictures, obtained either by stitching fluorescent microscopy images (Nikon Eclipse Ti) or by transmitted/reflected light macroscope (Nikon SMZ745T/Leica DMS1000) are used to measure the wavelength of the primary pattern, Supplementary Figures~\ref{S-fig:generations} and \ref{S-fig:patterns}. The secondary pattern wavelength is taken as half of the primary wavelength, according to Supplementary Figure~\ref{S-fig:generations}.
Data shown in Figure~\ref{fig:acidification}a and c are obtained with a SevenCompact pH-meter (Mettler Toledo) and an MCR 301 rheometer (Anton Paar). For every gel composition we monitor the gelation by small amplitude oscillations (strain 0.1\%, frequency \SI{1}{\hertz}) and estimate the flexural modulus $B$ from the maximum value of the shear modulus $G^\prime$ as $B=2G^\prime/(1-\nu)$ taking $\nu=0.3$ for the Poisson ratio, a value typical of a spongy network~\cite{Greaves2011}.
Permeability is measured using a protocol found in \cite{VanDijk1986}, see Supplementary Method and Supplementary Figure~\ref{S-fig:permeability}. The amount of free casein protein present in the gel during the GDL acidification process is measured by titration and UV-spectrometry (Ocean Optics) according to the protocol found in~\cite{Roefs1986} and detailed in Supplementary Method.
\end{materials}

\begin{acknowledgments}
Confocal experiments were conducted at SFR BioSciences Gerland - Lyon Sud (US8 / UMS3444).
The authors thank Jos\'e Bico, Arezeki Boudaoud, Cyprien Gay, L. Mahadevan and Olivier Pierre-Louis for theoretical insights at various stages of the research. 
Precious experimental ideas came from Catherine Barentin, Denis Bartolo, Elisabeth Bouchaud and Thibaut Divoux.
Alan Parker at Firmenich provided the casein and GDL. 
Special thanks to Madame Biot who made M.A. Biot work accessible through the \href{http://www.olemiss.edu/sciencenet/poronet/}{poronet} website. 
ML thanks the Region Rhône Alpes and the Programme d'Avenir Lyon - Saint Etienne (PALSE NoGELPo) for postoctoral grant. 
TG and MN acknowledge funding from the Agence Nationale de la Recherche (ANR-11-PDOC-027). 
SM and ML acknowledge funding from the European Research Council under the European Union's Seventh Framework Program (FP7/2007-2013) / ERC grant agreement No. 258803 and from Institut Universitaire de France.
\end{acknowledgments}

\bibliographystyle{pnas2009}
\bibliography{Yaourt}

\end{article}

%\paragraph*{Author Information} 
%The authors declare that they have no competing financial interests. 
%Correspondence and requests for materials should be addressed to TG (\href{mailto:thomas.gibaud@ens-lyon.fr}{thomas.gibaud@ens-lyon.fr}).
